---
title: "RBIF125_Final_Prior"
author: "Andrew Prior"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
    highlight: tango
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
# Use current directory (where this Rmd file is located)
# Works on both Windows and Linux
PROJECT_DIR <- getwd()
knitr::opts_knit$set(root.dir = PROJECT_DIR)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "results/figures/",
  cache = FALSE
)
```
# Long COVID Prediction from PBMC Transcriptomics

# 1. Setup & Configuration

## 1.1 Load Libraries

```{r load-libraries, message=FALSE}
library(DESeq2)
library(GEOquery)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(scales)
library(randomForest)
library(caret)
library(pROC)
library(e1071)
library(doParallel)
library(foreach)
```

## 1.2 Configuration

```{r config-params}

#set parameters for script
n_cores <- parallel::detectCores() - 1
n_cores <- max(1, n_cores)
registerDoParallel(cores = n_cores)
set.seed(42)

PADJ_THRESHOLD <- 0.05
LOG2FC_THRESHOLD <- 1.5
N_TOP_GENES <- 100
N_FOLDS <- 10

#create output folders
dir.create("results", showWarnings = FALSE)
dir.create("results/figures", showWarnings = FALSE, recursive = TRUE)
dir.create("results/tables", showWarnings = FALSE, recursive = TRUE)
dir.create("data/raw", showWarnings = FALSE, recursive = TRUE)
dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)
```

## 1.3 Custom Functions

```{r custom-functions}
# save plots to file
save_plot <- function(plot_obj, filename, width = 10, height = 7) {
  filepath <- file.path("results/figures", filename)
  ggsave(filepath, plot_obj, width = width, height = height, dpi = 300)
}
# save tables to file
save_table <- function(data_frame, filename) {
  filepath <- file.path("results/tables", filename)
  write.csv(data_frame, filepath, row.names = FALSE)
}
```

---

# 2. Data Loading

## 2.1 Retrieve/Load Metadata

```{r download-metadata}
#Retrieve GEO metadata

# set global download settings to prevent network timeouts
options(timeout = 600)
options(download.file.method = "libcurl")
options('download.file.method.GEOquery' = 'libcurl')

GEO_ACCESSION <- "GSE226260"
series_file <- file.path("data/raw", "GSE226260-GPL24676_series_matrix.txt.gz")

# load locally if file exists, otherwise download safely
if (file.exists(series_file)) {
  # skip passing getGPL when reading local files
  gse <- getGEO(filename = series_file)
} else {
  gse_list <- getGEO(
    GEO = GEO_ACCESSION, 
    GSEMatrix = TRUE, 
    destdir = "data/raw", 
    getGPL = FALSE
  )
  gse <- gse_list[[1]]
}

metadata <- pData(gse)
print(paste("Metadata dimensions:", nrow(metadata), "x", ncol(metadata)))
print("Metadata column names (first 10):")
print(head(colnames(metadata), 10))
print("Sample IDs (first 3):")
print(head(rownames(metadata), 3))
```

## 2.2 Download Expression Data

```{r download-expression}

#Retrieve GEO count data
combined_file <- "data/raw/GSE226260/GSE226260_combined.ExprData.updated.csv.gz"

if (!file.exists(combined_file)) {
  dir.create("data/raw/GSE226260", showWarnings = FALSE, recursive = TRUE)
  download.file(
    "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE226260&format=file&file=GSE226260%5Fcombined%2EExprData%2Eupdated%2Ecsv%2Egz",
    combined_file,
    mode = "wb"
  )
}

expr_data <- read.csv(gzfile(combined_file), check.names = FALSE)
gene_names <- make.unique(as.character(expr_data[[1]]), sep = "_")
rownames(expr_data) <- gene_names
expr_matrix <- as.matrix(expr_data[, -1])

print(paste("Expression dimensions:", nrow(expr_matrix), "genes x", ncol(expr_matrix), "samples"))
print(paste("Value range:", round(min(expr_matrix, na.rm = TRUE), 2), "to", round(max(expr_matrix, na.rm = TRUE), 2)))
print(expr_matrix[1:3, 1:3])
```

## 2.3 Align Samples

```{r align-samples}
#match metadata to counts data rows/columns
expr_gcids <- sub("_.*$", "", colnames(expr_matrix))
metadata_gcids <- metadata$`gcid:ch1`
common_gcids <- intersect(expr_gcids, metadata_gcids)

gcid_to_expr_col <- setNames(colnames(expr_matrix), expr_gcids)
expr_matrix <- expr_matrix[, gcid_to_expr_col[common_gcids]]
metadata <- metadata[match(common_gcids, metadata_gcids), ]

colnames(expr_matrix) <- common_gcids
rownames(metadata) <- common_gcids

print(paste("Aligned samples:", length(common_gcids)))
print(paste("Sample IDs match:", all(colnames(expr_matrix) == rownames(metadata))))
print("Aligned sample IDs (first 3):")
print(head(colnames(expr_matrix), 3))
```

## 2.4 Extract PASC Labels

```{r extract-pasc}
pasc_status <- metadata$`pasc status:ch1`
print("PASC status distribution:")
print(table(pasc_status, useNA = "ifany"))

metadata$severity_clean <- ifelse(pasc_status == "NOPASC", "Convalescent",
                                  ifelse(
                                    pasc_status == "PASC",
                                    "Long_COVID",
                                    NA)
                                  )
metadata$severity_clean <- factor(metadata$severity_clean, levels = c("Convalescent", "Long_COVID"))

has_label <- !is.na(metadata$severity_clean)
metadata <- metadata[has_label, ]
expr_matrix <- expr_matrix[, has_label]

print("Final dataset after filtering:")
print(table(metadata$severity_clean))
print(paste("Final dimensions:", nrow(expr_matrix), "genes x", ncol(expr_matrix), "samples"))
```

## 2.5 Prepare for DESeq2

```{r prepare-deseq}

expr_matrix_counts <- round(expr_matrix)

# remove genes with NA values
has_na <- rowSums(is.na(expr_matrix_counts)) > 0
print(paste("Genes with NA values:", sum(has_na)))
expr_matrix_counts <- expr_matrix_counts[!has_na, ]
print(paste("After removing NA genes:", nrow(expr_matrix_counts), "genes x", ncol(expr_matrix_counts), "samples"))

# filter out low-count genes
keep_genes <- rowSums(expr_matrix_counts) >= 10
print(paste("Genes passing filter (>= 10 total counts):", sum(keep_genes)))
expr_matrix_counts <- expr_matrix_counts[keep_genes, ]
print(paste("Final count matrix:", nrow(expr_matrix_counts), "genes x", ncol(expr_matrix_counts), "samples"))

coldata <- data.frame(
  sample_id = rownames(metadata),
  severity = metadata$severity_clean,
  row.names = rownames(metadata)
)
print("Sample metadata for DESeq2 (first 3):")
print(head(coldata, 3))

#Create DEseq2 object
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = expr_matrix_counts,
  colData = coldata,
  design = ~ severity
)

print(paste("DESeq2 object:", nrow(dds), "genes x", ncol(dds), "samples"))
print("Sample group distribution:")
print(table(colData(dds)$severity))
```

---

# 3. Quality Control

## 3.1 Library Size Distribution

```{r library-sizes, fig.width=10, fig.height=6}
# Calculate total counts per sample
library_sizes <- colSums(expr_matrix_counts)

lib_size_df <- data.frame(
  sample = colnames(expr_matrix_counts),
  library_size = library_sizes,
  severity = metadata$severity_clean,
  row.names = NULL
)

print("Library size summary:")
print(summary(library_sizes))
print("Library sizes by group:")
print(tapply(lib_size_df$library_size, lib_size_df$severity, summary))

# Plot library size distribution
library_size_plot <- ggplot(lib_size_df, aes(x = severity, y = library_size, fill = severity)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("Convalescent" = "blue", "Long_COVID" = "red")) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Library Size Distribution by Group",
    subtitle = "Log10 scale",
    x = "Group",
    y = "Library Size (Total Counts)"
  ) 

print(library_size_plot)
ggsave("results/figures/library_sizes.png", library_size_plot, width = 10, height = 6)
```

## 3.2 Principal Component Analysis

```{r pca-analysis, fig.width=10, fig.height=7}
# Filter genes with zero variance before PCA
gene_vars <- apply(expr_matrix_counts, 1, var, na.rm = TRUE)
genes_with_var <- gene_vars > 0 & !is.na(gene_vars)
expr_matrix_filtered <- expr_matrix_counts[genes_with_var, ]

print(paste("Genes before variance filtering:", nrow(expr_matrix_counts)))
print(paste("Genes with variance > 0:", nrow(expr_matrix_filtered)))
print(paste("Removed constant genes:", sum(!genes_with_var)))

# Perform PCA 
pca_result <- prcomp(t(expr_matrix_filtered), scale. = TRUE, center = TRUE)

# Extract PC values
pca_scores <- as.data.frame(pca_result$x[, 1:10])
pca_scores$sample <- rownames(pca_scores)
pca_scores$severity <- metadata$severity_clean

# Calculate variance explained
var_explained <- round(100 * pca_result$sdev^2 / sum(pca_result$sdev^2), 1)

print("Variance explained by first 5 PCs:")
print(var_explained[1:5])

# Plot PC1 vs PC2
pca_plot <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = severity, shape = severity)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("Convalescent" = "blue", "Long_COVID" = "red")) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    title = "Principal Component Analysis",
    subtitle = "Sample clustering by group",
    x = paste0("PC1 (", var_explained[1], "% variance)"),
    y = paste0("PC2 (", var_explained[2], "% variance)"),
    color = "Group",
    shape = "Group"
  ) +
  theme(legend.position = "right")

print(pca_plot)
ggsave("results/figures/pca_plot.png", pca_plot, width = 10, height = 7)
print("Saved: results/figures/pca_plot.png")
```

## 3.3 Gene Detection Rates

```{r gene-detection, fig.width=10, fig.height=6}
# Calculate genes detected per sample (count > 0)
genes_detected <- colSums(expr_matrix_counts > 0)

detection_df <- data.frame(
  sample = names(genes_detected),
  genes_detected = genes_detected,
  severity = metadata$severity_clean,
  row.names = NULL
)

print("Gene detection summary:")
print(summary(genes_detected))
print("Gene detection by group:")
print(tapply(detection_df$genes_detected, detection_df$severity, summary))

# plot gene detection rates for each group
gene_detection_plot <- ggplot(detection_df, aes(x = severity, y = genes_detected, fill = severity)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16) +
  geom_jitter(width = 0.2, size = 1.5) +
  scale_fill_manual(values = c("Convalescent" = "blue", "Long_COVID" = "red")) +
  labs(
    title = "Gene Detection Rate by Group",
    x = "Group",
    y = "Number of Genes Detected"
  ) +
  theme(legend.position = "none")

print(gene_detection_plot)
ggsave("results/figures/gene_detection.png", gene_detection_plot, width = 10, height = 6)
```

---

# 4. Differential Expression Analysis

## 4.1 Run DESeq2

```{r run-deseq}
print("Running DESeq2 pipeline (normalization, dispersion, testing)...")
dds <- DESeq(dds)
print("DESeq2 analysis complete")
```

## 4.2 Extract Results

```{r extract-de-results}
res_longcovid_vs_conv <- results(
  dds,
  contrast = c("severity", "Long_COVID", "Convalescent"),
  alpha = 0.05
)

res_df <- as.data.frame(res_longcovid_vs_conv)
res_df$gene_id <- rownames(res_df)
res_df <- res_df[order(res_df$padj), ]

res_df$significant <- res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1.5
res_df$direction <- ifelse(
  res_df$log2FoldChange > 1.5 & res_df$padj < 0.05,
  "Up",
  ifelse(
    res_df$log2FoldChange < -1.5 & res_df$padj < 0.05,
    "Down",
    "NS"
  )
)

print("Differential Expression Summary:")
print(paste("Total genes tested:", nrow(res_df)))
print(paste("Significant genes (padj < 0.05):", sum(res_df$padj < 0.05, na.rm = TRUE)))
print(paste("Upregulated in Long COVID (log2FC > 1.5, padj < 0.05):", sum(res_df$direction == "Up", na.rm = TRUE)))
print(paste("Downregulated in Long COVID (log2FC < -1.5, padj < 0.05):", sum(res_df$direction == "Down", na.rm = TRUE)))

print("Top 10 significant genes:")
sig_genes_top10 <- res_df[res_df$significant == TRUE, c("gene_id", "log2FoldChange", "padj")]
print(head(sig_genes_top10, 10))

write.csv(res_df, "results/tables/de_results_longcovid_vs_convalescent.csv", row.names = FALSE)
print("Saved: results/tables/de_results_longcovid_vs_convalescent.csv")

#Annotate top DE genes for display (Ensembl ID -> Symbol + Name)
top_de_genes <- res_df[res_df$significant == TRUE, ]
top_de_genes <- top_de_genes[order(top_de_genes$padj), ]

if (nrow(top_de_genes) > 0) {
  # add Ensembl IDs to gene symbols and names
  gene_annotation <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = top_de_genes$gene_id,
    columns = c("SYMBOL", "GENENAME"),
    keytype = "ENSEMBL"
  )

  # Merge annotations (keep all Ensembl IDs, allow duplicates if genes map to multiple symbols)
  top_de_annotated <- merge(
    top_de_genes,
    gene_annotation,
    by.x = "gene_id",
    by.y = "ENSEMBL",
    all.x = TRUE
  )
  top_de_annotated <- top_de_annotated[order(top_de_annotated$padj), ]

  # save annotated top DE genes for human review
  write.csv(top_de_annotated, "results/tables/top_de_genes_annotated.csv", row.names = FALSE)

  print("Top 20 DE genes with gene symbols:")
  print(top_de_annotated[1:min(20, nrow(top_de_annotated)), c("gene_id", "SYMBOL", "GENENAME", "log2FoldChange", "padj")])
}
```

---

# 5. Visualization

## 5.1 Volcano Plot

```{r volcano-plot, fig.width=10, fig.height=8}
volcano_data <- res_df[!is.na(res_df$padj), ]
volcano_data$neg_log10_padj <- -log10(volcano_data$padj)
volcano_data$color_group <- ifelse(volcano_data$direction == "Up", "Upregulated in Long COVID",
                                   ifelse(
                                     volcano_data$direction == "Down",
                                     "Downregulated in Long COVID",
                                     "Not Significant")
                                   )

volcano_plot <- ggplot(volcano_data, aes(x = log2FoldChange, y = neg_log10_padj, color = color_group)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "black") +
  scale_color_manual(
    values = c("Upregulated in Long COVID" = "red",
               "Downregulated in Long COVID" = "blue",
               "Not Significant" = "black"),
    name = "Gene Status"
  ) +
  labs(
    title = "Volcano Plot: Long COVID vs Convalescent",
    x = "Log2 Fold Change",
    y = "-Log10(Adjusted P-value)"
  ) 

print(volcano_plot)
ggsave("results/figures/volcano_plot_longcovid.png", volcano_plot, width = 10, height = 8)
```

---

# 6. Variance Stabilizing Transformation

```{r vst-transform}
# VST stabilizes variance across the mean, making the data suitable for PCA and Random Forest

vst_data <- vst(dds, blind = FALSE)
vst_matrix <- assay(vst_data)
print(paste("VST matrix dimensions:", nrow(vst_matrix), "x", ncol(vst_matrix)))
print(paste("VST value range:", round(min(vst_matrix), 2), "to", round(max(vst_matrix), 2)))
print("VST data preview (first 3 genes, first 3 samples):")
print(vst_matrix[1:3, 1:3])
```

---

# 7. Machine Learning: Feature Selection

```{r feature-selection}
# Select top differentially expressed genes as features for Random Forest

res_df_clean <- res_df[!is.na(res_df$padj), ]
res_df_sorted <- res_df_clean[order(res_df_clean$padj), ]
feature_genes <- head(res_df_sorted$gene_id, N_TOP_GENES)

print("Top 10 feature genes:")
print(head(feature_genes, 10))

# Transpose to samples as rows, features as columns for input to Random Forest
feature_matrix <- t(vst_matrix[feature_genes, ])
ml_labels <- metadata$severity_clean

ml_data <- as.data.frame(feature_matrix)
ml_data$severity <- ml_labels

print(paste("Random Forest dataset:", nrow(ml_data), "samples x", ncol(ml_data)-1, "features"))
print("Class distribution in dataset:")
print(table(ml_data$severity))
print("ML data preview (first 3 samples, first 5 features + label):")
print(ml_data[1:3, c(1:5, ncol(ml_data))])
```

---

# 8. Random Forest Classifier

## 8.1 Setup Cross-Validation

```{r rf-cv-setup}
# K-fold CV: Data split into N_FOLDS folds, train on K-1, test on 1, repeat K times
train_control <- trainControl(
  method = "cv",
  number = N_FOLDS,
  savePredictions = "final",
  classProbs = TRUE,
  allowParallel = TRUE
)

# mtry = number of features randomly sampled at each split
rf_grid <- expand.grid(
  mtry = round(c(sqrt(N_TOP_GENES), N_TOP_GENES/3, N_TOP_GENES/2))
)

print(rf_grid)
print(paste("Total models to train:", nrow(rf_grid), "x", N_FOLDS, "folds =", nrow(rf_grid) * N_FOLDS, "fits"))
```

## 8.2 Train Random Forest

```{r rf-train}

print(paste("Dataset:", nrow(ml_data), "samples x", ncol(ml_data)-1, "features"))
print(paste("Classes:", paste(levels(ml_data$severity), collapse = ", ")))

#set for reproducible results
set.seed(3)

rf_model <- train(
  severity ~ .,
  data = ml_data,
  method = "rf",
  trControl = train_control,
  tuneGrid = rf_grid,
  ntree = 500,
  importance = TRUE
)

print(rf_model)
```

---

# 9. Model Evaluation

## 9.1 Confusion Matrix

```{r confusion-matrix, fig.width=8, fig.height=7}
predictions <- rf_model$pred[rf_model$pred$mtry == rf_model$bestTune$mtry, ]

cm <- confusionMatrix(predictions$pred, predictions$obs)
print(cm)

cm_table <- as.data.frame(cm$table)
write.csv(cm_table, "results/tables/confusion_matrix.csv", row.names = FALSE)

cm_plot <- ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Confusion Matrix", x = "Actual", y = "Predicted") 

print(cm_plot)
ggsave("results/figures/confusion_matrix.png", cm_plot, width = 8, height = 7)
```

## 9.2 Feature Importance

```{r feature-importance, fig.width=10, fig.height=8}
# varImp returns class-specific scores for binary classification, 
# use first column
importance_scores <- varImp(rf_model)$importance
importance_df <- data.frame(
  gene = rownames(importance_scores),
  importance = importance_scores[, 1]  # Use first column for binary classification
)
importance_df <- importance_df[order(-importance_df$importance), ]
importance_df <- head(importance_df, 20)

print("Top 20 most important features:")
print(importance_df)

# For feature importance genes, add symbols and names
feature_annotation <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = importance_df$gene,
  columns = c("SYMBOL", "GENENAME"),
  keytype = "ENSEMBL"
)

# Merge annotations for combined result
importance_df_annotated <- merge(
  importance_df,
  feature_annotation,
  by.x = "gene",
  by.y = "ENSEMBL",
  all.x = TRUE
)
importance_df_annotated <- importance_df_annotated[order(-importance_df_annotated$importance), ]

# Save annotated feature importance csv file
write.csv(importance_df_annotated, "results/tables/feature_importance_annotated.csv", row.names = FALSE)
print("Saved: results/tables/feature_importance_annotated.csv")

print("Top 20 most important features:")
print(importance_df_annotated[, c("gene", "SYMBOL", "GENENAME", "importance")])

importance_plot <- ggplot(importance_df, aes(x = reorder(gene, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  labs(title = "Top 20 Most Important Features", x = "Gene", y = "Importance Score") 

print(importance_plot)
ggsave("results/figures/feature_importance.png", importance_plot, width = 10, height = 8)
```

## 9.3 Summary

```{r summary}
cat("Overall Accuracy:", round(cm$overall["Accuracy"], 3), "\n")
cat("Kappa:", round(cm$overall["Kappa"], 3), "\n")
cat("Best mtry:", rf_model$bestTune$mtry, "\n")
```

---

# 10. Session Info

```{r session-info}
sessionInfo()
```


